---
title: ""
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(RColorBrewer)
library(googlesheets4)
library(plotly)
library(dplyr)
library(rsconnect)
```

</style>

<div class="row" style="padding-top: 30px;">
<div class="col-sm-10">
<div align="justify">

## Current trends

### Overall GHG emissions

Between 2016 and 2020 GHG emission from canton Sarajevo grew at a rate of about 2.1% a year - see Figure 1 grey bars. The dip in 2020 motivated by COVID-19 lock-down is likely to be short-lived and a rebound on emissions in 2021 and 2022 is expected in line with [global trends](https://www.nature.com/articles/d41586-021-03036-x). Assuming a relative rebound of emissions for 2021 similar to that observed for the [EU](https://www.nature.com/articles/d41586-021-03036-x) implies an estimate of about 2300 Thousand tons CO~2eq~ for canton Sarajevo - see Figure 1 blue bars. Such a development would further place the current emission trajectory at odds with that necessary for a 80% cut in emission by 2050 - see Figure 1 grey bars. On a per-capita basis the emissions in canton Sarajevo are estimated at circa 5.1 tons CO~2eq~ for the year 2021. Details about the methodology, scope and data sources used can be found in the GHG emission inventory tab.

```{r monitor,echo=FALSE, fig.cap="Figure 1 - GHG emissions in canton Sarajevo"}

emi.monitor<-cbind.data.frame("Year" = c(2016:2050),
                              "GHG"=c(2000.68,2026.70,2116.71,2276.25,2163.30,
                                      2300,2300,
                                      seq(2163.30,457.6, length.out=28)))

p<-plot_ly(emi.monitor, x =~Year, y=~GHG, 
        type = "bar" ,  #specifies the chart is a scatter plot
        #color= c(rep("blue",5), rep("red",30)),
        colors=colorRamp(c( "skyblue","grey60","palegreen4")),
        color= c(rep("Observed",5),rep("Assuming 6.5% rebound",2),rep("Trajectory for 80% reductions in 2050",28)),
        width = 700,
        height = 250)

p <- p %>% layout(title = '',
         xaxis = list(title = 'Year',
                      zeroline = TRUE,
                      range = c(2015,2051)),
         yaxis = list(title = 'Thousand tons (CO2e)',
                      range = c(0,2700)))
p

```

### Sector decomposition

In terms of sector breakdown, residential buildings and road transport dominate the share of total emissions in the canton - see Figure 2. Combined, these two sectors were associated with over 60% of total emissions in the year 2020. Non-residential buildings associated with services and commerce make up circa 13% of emissions while the industry sector is associated with 15% of total GHG emissions. Emissions from electricity are accounted in each sector respectively. How sector-specific emissions evolved post-pandemic is unclear but judging from other [monitoring services](https://carbonmonitor.org/variation) transport, power and industry should lead the way. Note that the sector breakdown presented in Figure 2 tries to be as close as possible to that required by the [Covenant of Mayors](https://www.globalcovenantofmayors.org/) (CoM) but other allocations are possible.

```{r sector, echo=FALSE, fig.cap="Figure 2 - GHG emissions in the year 2020"}

p2 <- plot_ly(cbind.data.frame("Sector"=c("Residential buildings","Tertiatry buildings","District heating",
                                          "Public lighting","Municipal buildings","Road transport","Industry",
                                          "Waste"), "GHG"=c(562.78,231.05,115.15,19.09,10.93,572.07,270.49,31.32)),
              labels = ~Sector, values = ~GHG, type = 'pie',width = 600, height = 250)
p2 <- p2 %>% layout(title = '',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))
p2

```
Sectors exhibit marked differences regarding past trends of GHG emissions - see Table 1. Disregarding the year 2020 due to the economic and behavior distortions introduced by the global pandemic, one can observe a sharp increase of yearly emissions from the household sector (including electricity and district heating) in the order of 4.5%/year. Road transportation has increased about 1.6%/year between 2016 and 2019 while the increase in municipal services (which includes gas consumption in municipal buildings and public lighting) has been moderate and below 1%/year.

```{r, echo=FALSE}
sector.t<-c("Households","Tertiatry buildings","Road transport","Industry","Municipal services")
sector.tv<-c(4.5,1.6,1.4,1.3,0.8)
  
taba<-cbind.data.frame("Sector"=sector.t,"Average % change"=sector.tv)
knitr::kable(taba, "simple", caption = "Table 1 - Average change for the 2016-2019 period")

```

***